Sentence generation with RTGs

### 2009-06-17 trunk/Papers/crisp-rtg-09/crisp-rtg-09.tex
ported npc proof and crisp encoding

\usepackage{url}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsfonts,amsthm}

@@

@@

\section{The generation problem of regular tree grammars}

@@

\section{RTG generation as planning}

\section{NP-completeness of the RTG generation problem}

>>>>

\usepackage{url}
\usepackage{graphicx}
\usepackage[round]{natbib}
\usepackage{amsfonts,amsthm}

@@

\newcommand{\actionname}[3]{\mathbf{\mathtt{#1}\mbox{-}\mathtt{#2}\mbox{-}\mathtt{#3}}}
\newcommand{\action}[4]{\textbf{Action \texttt{#1}$(#2)$.}\\
\strut\quad   Precond:$\;$ \parbox[t]{12cm}{\ensuremath{#3}}\\
\strut\quad   Effect:$\;$ \parbox[t]{12cm}{\ensuremath{#4}}}
\newcommand{\actionc}[4]{\textbf{Action \texttt{#1}$(#2)$.}
Precond:$\;$ \parbox[t]{3cm}{\ensuremath{#3}}\\
\strut\quad   Effect:$\;$ \parbox[t]{12cm}{\ensuremath{#4}}}

\newcommand{\open}{\ensuremath{\mathsf{open}}}
\newcommand{\gensym}{\ensuremath{\mathsf{gensym}()}}

@@

\section{The generation problem of regular tree grammars}
\label{sec:gener-probl-regul}

@@

\section{RTG generation as planning} 
\label{sec:rtg-generation-as-planning}

It is straightforward to extend the encoding of TAG sentence
generation to planning \citep{KolSto07} to regular tree grammars.
Notice that this makes the \citet{KolSto07} encoding a special case of
the algorithm presented here, as the derivation trees of a given TAG
grammar form regular tree languages
\citep{schmitz08:_featur_in_tag_deriv_trees} and can therefore be
described by RTGs.\footnote{This is not strictly speaking true,
  because we have not shown how to represent the pragmatic conditions
  and semantic requirements in an RGG, in order to keep the
  presentation simple; but it is obvious how these could be added.}

Let $l$ be a lexicon entry with production rule $A \rightarrow
\sigma(B_1 \Neq r_1, \ldots, B_n \Neq r_n)$.  Then the grammaticality
condition is encoded by the precondition and effects of a planning
operator as follows:\footnote{I write $r(u)$ to indicate the
  construction of a new symbol representing the $r$-child of $u$. I
  take $\self(u) = u$ as a special case. This cannot be expressed in
  PDDL, but can be encoded with the same planning-step-counting
  construction as in the ACL paper.}

\strut\\
\action{$\sigma$/$L$}{u}
{\open(u,A)}
{\neg \open(u,A), \open(r_1(u),,A_1), \ldots, \open(r_n(u),A_n)}\\

Notice that I write $\open$ to encode the unprocessed nodes in the
tree we are constructing, rather than $\mathsf{subst}$ and
$\mathsf{adj}$ as in the paper. Those were TAG-specific predicates,
encoding TAG's substitution and adjunction operation, and are no
longer necessary here. Effectively, what happens here is that the plan
encodes a derivation of the underlying RTG grammar; an atom
$\open(u,A)$ in the current planning state encodes the fact that the
current partial RTG derivation contains a leaf $u$ that is labeled
with the nonterminal symbol $A$ (and still needs to be expanded in the
future).

The translation of the semantic and reference conditions then proceeds
essentially as in the paper.  However, one crucial difference is that
we no longer distinguish between substitution and adjunction, and thus
it is not completely obvious where distractors for new referring
expressions get introduced. This used to happen exactly at the
substitution nodes. Now we can get the same effect by introducing
distractors only for those nonterminal occurrences on the right-hand
side that are labeled with a semantic role that is not $\self$.

Note that the encoding of RGG generation as a planning problem is not
entirely correct.  The difference is that in the problem as defined in
Section~\ref{sec:gener-probl-regul}, we took $a$ to be a possible
distractor for a new referring expression if there was any tuple
$(a_1,\ldots,a,\ldots,a_n)$ that satisfied the semantic condition of
the lexicon entry that introduced the RE.  In the encoding into
planning in the ACL paper, however, we only accept $a$ as a possible
distractor if $a_1,\ldots,a_n$ are the intended referents for the
other semantic roles in the lexicon entry. If we continue using this
encoding here, we end up with a planning problem that generates too
few distractors, and therefore can compute referring expressions that
are not unique. This is something that needs to be fixed at some
point, but for now the planning encoding seems to work pretty well in
practice.

\section{NP-completeness of the RTG generation problem}
\label{sec:np-completeness}

Finally, we can show that the sentence generation problem of RGGs is
NP-complete. This means that it is combinatorially hard, and hence
solving it with a worst-case exponential-time algorithm like a planner
is appropriate.

The NP-hardness proof below works by reducing the HAMILTONIAN-CYCLE
problem \cite{garey79:_comput_and_intrac} to the sentence generation
problem. HAMILTONIAN-CYCLE is the (NP-complete) problem of deciding
whether a directed graph has a Hamiltonian cycle.  A Hamiltonian cycle
is a path $(u_1,u_2,\ldots,u_n,u_1)$ in the graph such that the graph
contains an edge $(u_i,u_{i+1})$ for every $1 \leq i <n$ and the edge
$(u_n,u_1)$ -- i.e., it is a cycle --, and such that the cycle visits
every node in the graph exactly once. In the example graph in
Fig.~\ref{fig:npc}a, $(1,3,2,1)$ is a Hamiltonian cycle; $(1,2,1)$ is
a cycle that is not Hamiltonian because it doesn't visit 3.

\begin{prop} \label{prop:npc} The sentence generation problem of
  regular generation grammars is NP-complete.
\end{prop}
\begin{proof}
  It is obvious that the problem is in NP; once we have
  nondeterministically guessed a candidate tree, we can test whether
  it satisfies all conditions in polynomial time.

  For the NP-hardness, let's say we have a graph $G =
  (\{1,\ldots,n\},E)$ with $m$ edges of the form $(i,k)$ of which we
  want to check whether it has a Hamiltonian cycle.  We construct a
  regular generation grammar $G_G$ with nonterminals $N = \{S,
  A_{11}\} \cup \{A_{ik} \;|\; 2 \leq i \leq n+1, 1 \leq k \leq n\}$,
  start symbol $S$, terminal alphabet $\Sigma=\{f,a\}$, role alphabet
  $R = \{\mathsf{r}, \self\}$, and the following lexicon entries:

  $$\begin{array}{l}
    S \rightarrow f(A_{11} \Neq \mathsf{r}) \\
    \mathsf{sem} = \{ \mathsf{start}(\self,\mathsf{r})\} \\
    \\
    A_{ij} \rightarrow f(A_{(i+1)k} \Neq \self) \quad \mbox{for all
      $(j,k) \in E$ and $1 \leq i \leq n$} \\ 
    \mathsf{sem} = \{ \mathsf{neq}_j(\self) \} \\
    \\
    A_{(n+1)1} \rightarrow a \\
    \mathsf{sem} = \emptyset
  \end{array}
  $$

  Now the target referent $e$ and the communicative goal $\emptyset$
  have a successful derivation given $G_G$ and the knowledge bases
  $SKB = HKB = \{\mathsf{start}(e,1)\} \cup \{ \mathsf{neq}_i(k) \;|\;
  1 \leq i,k \leq n \;\mbox{and}\; i \neq k \}$ iff $G$ has a
  Hamiltonian cycle.  To see this, notice that any tree generated by
  $G_G$ must be a linear tree with nonterminal symbols $S, A_{11},
  A_{2u_2}, \ldots, A_{(n+1)1}$ from top to bottom, and for any two
  subsequent nonterminals $A_{ij}$ and $A_{(i+1)k}$ in this sequence,
  there must be an edge $(j,k)$ in the graph.  So the tree encodes a
  cycle of length $n$.  Now, if any node in the graph, say $j$, was
  not visited on this cycle, then the semantics in the subtree of the
  referring expression with nonterminal $A_{11}$ does not contain the
  atom $\mathsf{neq}_j(1)$.  In other words, the tree does not contain
  sufficient information to distinguish 1 (the target referent of the
  referring expression) from $j$, so the referring expression is not
  unique, and the derivation is not successful.  Hence, the cycle
  visits every node exactly once and is therefore Hamiltonian.
\end{proof}

\begin{figure}
  \centering
  \begin{tabular}{cp{2cm}c}
  \includegraphics[scale=0.5]{pic-npc-graph}
  &&
  \includegraphics[scale=0.5]{pic-npc} \\
  (a) && (b)
  \end{tabular}
  \caption{(a) A graph with a Hamiltonian cycle, and (b) a tree
    encoding this cycle according to the construction in
    Prop.~\ref{prop:npc}.} 
  \label{fig:npc}
\end{figure}

An example of this construction (i.e., a graph together with the tree
that encodes its Hamiltonian cycle and that is generated by the RGG)
is shown in Fig.~\ref{fig:npc}. Notice that nothing in this
construction particularly hinges in the generalization of the TAG
generation problem to RGGs. The proof can thus be ported directly to a
proof of the NP-completeness of the TAG sentence generation problem.
